Contracts for unions and intersections are hard for Nickel

### 2021-06-08 paper.tex
Continue reworking section 5 (Keil&Thiemann)

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let C = contracts.fromPred (fun f =>
  f 0 == 0) in
let g | (String -> String) @& C
      = fun x => x
\end{lstlisting}
\caption{Intersection and user defined contracts}
\label{fig:inter-flat-contracts}
\end{figure}

@@

@@

The existing work that previously researched and implemented
union and intersection
as dynamic checks have, in our eyes, different shortcomings
that makes them suboptimal solutions.
In this section, we informally present those issues,
starting with the foundational paper by Keil and Thiemann
\cite{KeilThiemannUnionIntersection}, then following to a recent
work by Williams, Morris, and Wadler \cite{RootCauseOfBlame}, and
finally going over some similar contracts available in the Racket
programming language.

\subsection{Keil and Thiemann}
\label{sec:keil-thiemann}
\unsure{Subsection title?}
Keil and Thiemann noticed that, since union contracts should share information between
different application contexts, while intersection contracts should behave
independently between each application context, the system becomes much
simpler if every contract gets transformed into a union of intersections
of other contracts, by using the common De Morgan laws,
and then eagerly opening up unions, but postponing intersection
contracts as much as possible.
\todo{This assumes the difficulty has been previously explained}

For instance, if someone were to wrap a value \nickel{x}
in a contract as shown on the Original code snippet on Figure
\ref{fig:kt-contract-transf}, then, before actually executing that
contract, it would get transformed, and intersections would
be pushed inside the unions, as can be seen on the Transformed
code snippet on the same figure.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
x | A @& (B @| C)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Transformed]
x | (A @& B) | (A @& C)
\end{lstlisting}
\caption{Contract transformation performed by the Keil and
Thiemann work}
\label{fig:kt-contract-transf}
\end{figure}

This straightforward idea makes composing the Keil and Thiemann
solution with other existing systems more difficult; in the words
of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
``(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
special cases can hinder composition.''\cite{RootCauseOfBlame}.
\info{(Yann) Does
it also make it less efficient? If $(A \cap B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
And it also could harm performance, or at least complicate an implementation,
since some sharing mechanism should be used to avoid checking the same contract,
over the same value, multiple times, exemplified by contract \nickel{A} on
Figure \ref{fig:kt-contract-transf}, that gets duplicated.
\unsure{Arnaud says: this whole
section is too abstract give concrete code and explain what goes
wrong. It's fine to give code in Nickel syntax}

Another difficult issue Keil and Thiemann were able to solve, is the
fact that user defined contracts and intersection do not usually play nice
together, as shown on Section \ref{sec:flat-and-inter}.
They provide two different sets of rules for their calculus,
the first one is non deterministic, and its purpose is to provide
semantics that are simple to understand.
With these rules, the interaction between user defined contracts and
intersections becomes much simpler, since they are able to check
both sides of an intersection on separate threads.
This allows them to be sure that contracts from one branch of the intersection
won't get blamed while executing the other branch, and vice versa.

As an example, consider the programs shown on Figure
\ref{fig:kt-non-deterministic}, if the Original program
gets executed, it will spawn two instances of said program,
one with the first branch of the intersection (\nickel{3 | Num}),
and the other one with the second branch (\nickel{3 | Positive}),
and these will get joined once they reach an equal point, in this case
it will be \nickel{3}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
3 | Num @& Positive
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=First Program]
3 | Num
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Second Program]
3 | Positive
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Joined Program]
3
\end{lstlisting}
\caption{Non deterministic intersection contract opening}
\label{fig:kt-non-deterministic}
\end{figure}

The second set of rules is deterministic, and it is intended
to be simpler to implement.
However, this second set needs to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.
\footnote{You can check more on this topic on Section 5.1 of
\cite{KeilThiemannUnionIntersection}}
For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
when using the value \nickel{(fun x => x) | Str -> Str)}
inside the execution of contract \nickel{C}, and applying it to \nickel{0},
the contract runtime would check the context of execution and it would
deduce that contract \nickel{Str -> Str} does not apply in this case,
skipping it.

We see two problems here, a minor one is the fact that the two
different semantics defined for the calculus are really far apart
one from the other, both on the strategies used for contract resolution,
as well as on the complexity levels.
\resolved{(Yann) Is it really a
    problem per se? It is common thing to do (a nice declarative system
    for proofs and an algorithmic one for implementation, proved equivalent).
    Maybe the point is that the algorithmic system is way more complex than the
declarative one
(Teo) reworded}.
A second, major one,
is the dependency added over the evaluation context, since, again,
it makes composing this, with other features desired on programming languages,
difficult, or even impossible.
\resolved{(Yann) I think we need to substantiate this claim. If this
context dependency explains why CSE is invalid, maybe have a little example?
(Teo) I removed it, I don't think there's an example, mainly since
this context dependency is not part of the language. I do think it complicates
implementation of said feature}
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}

>>>>

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let c = contracts.frompred (fun f =>
  f 0 == 0) in
let g | (string -> string) @& c
      = fun x => x
\end{lstlisting}
\caption{intersection and user defined contracts}
\label{fig:inter-flat-contracts}
\end{figure}

@@

In this section, we review some shortcomings of existing work on union and
intersection contracts. We start with the work of Keil and
Thiemann~\cite{KeilThiemannUnionIntersection} which is the first to propose a
convincing semantics for union and intersection contracts. We follow with the
work Williams, Morris, and Wadler~\cite{RootCauseOfBlame} which builds
upon~\cite{KeilThiemannUnionIntersection} and introduces a simpler, more
uniform blame resolution.

% The existing work that previously researched and implemented
% union and intersection
% as dynamic checks have, in our eyes, different shortcomings
% that makes them suboptimal solutions.
% In this section, we informally present those issues,
% starting with the foundational paper by Keil and Thiemann
% \cite{KeilThiemannUnionIntersection}, then following to a recent
% work by Williams, Morris, and Wadler \cite{RootCauseOfBlame}, and
% finally going over some similar contracts available in the Racket
% programming language.

\subsection{Keil and Thiemann}
\label{sec:keil-thiemann}
\unsure{Subsection title?}

In Section~\ref{sec:issues-sem}, we've seen that different calls to a function
with an union contract must share information: the behavior of one call is
influenced by the previous ones, as the function must pick one component of the
union to satisfy across all usages. Conversely, following the semantics of
overloading, each application of a function with an intersection contract can
select a different branch and is thus independent from the others. A general
contract composed of nested unions, intersections and higher-order contracts
appears to require complex book-keeping in order to correctly raise blame.

An aspect of the approach of~\cite{KeilThiemannUnionIntersection} is to rewrite
these nested contracts using the De Morgan law of
Figure~\ref{fig:de-morgan-law}, eagerly opening up unions while postponing
intersection as much as possible.  \todo{This assumes the difficulty has been
previously explained} This transformation flattens arbitrarily nested unions and
intersections to a union of intersections, making blame resolution more
manageable.\unsure{(Yann) find better (more precise) wording}

\begin{figure}[h]
\begin{lstlisting}[frame=none, numbers=none, language=nickel]
A @& (B @| C) (*@ $\simeq$ @*) (A @& B) @| (A @& C)
\end{lstlisting}
\caption{De Morgan law for contracts}
\label{fig:de-morgan-law}
\end{figure}

To implement this solution, \cite{KeilThiemannUnionIntersection} resorts to
specific reduction rules for unions and intersections which perform this
rewriting on the fly. The result is that the reduction rules sometimes feel
ad-hoc, and make us question the composability and extensibility of the system.
In the words of Williams, Morris, and Wadler (see
Section~\ref{sec:will-morr-wadl}): ``(\ldots) the monitoring semantics for
contracts of intersection and union types given by Keil and Thiemann are not
uniform. (\ldots) If uniformity helps composition, then special cases can hinder
composition.'' \cite{RootCauseOfBlame}. Another less important consequence is
that in Figure~\ref{fig:de-morgan-law}, the contract \nickel{A} is duplicated,
potentially causing additional work. If \nickel{B} fails, \nickel{A} will be
re-evaluated in each branch of the right hand side of
Figure~\ref{fig:de-morgan-law} in the absence of an additional sharing
mechanism.
\resolved{(Yann) Does
it also make it less efficient? If $(A \cap B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
\unsure{Arnaud says: this whole
section is too abstract give concrete code and explain what goes
wrong. It's fine to give code in Nickel syntax}

One important advantage of~\cite{KeilThiemannUnionIntersection} is to handle
user-defined contracts. We explained in Section~\ref{sec:issues-sem} why a
direct approach for higher-order intersection contracts (as done
in~\cite{RootCauseOfBlame}) is not trivial to adapt for user-defined contracts.
\cite{KeilThiemannUnionIntersection} provides two operational semantics of the
contract system: a non-deterministic one, more intuitive and proof-friendly, and
a deterministic one for actual implementation. The non-deterministic semantics
cleanly handle intersections with user-defined contracts: as each component is
checked in a separate thread, there is no direct interaction.

Let us recast the problematic example of Figure~\ref{fig:inter-flat-contracts}
in Figure~\ref{fig:inter-flat-contracts-2}). The execution now splits the
intersection into two sub-contracts that are checked separately. If one
component fails, a flag in a global state. In particular, the two threads don't
interact directly, and don't raise blame by themselves, leading this example run
as expected.

\begin{figure}[h]
\begin{lstlisting}[language=nickel,title={Original program}]
let C = contracts.fromPred (fun f =>
  f 0 == 0) in
let g | (String -> String) @& C
      = fun x => x
in g 0
\end{lstlisting}
\begin{lstlisting}[numbers=none,title={First component}]
g 0 | Str -> Str
\end{lstlisting}
\begin{lstlisting}[numbers=none,title={Second component}]
g 0 | C
\end{lstlisting}
\caption{Intersection and user defined contracts}
\label{fig:inter-flat-contracts-2}
\end{figure}

In return, the complexity is offloaded to a separate procedure that must decide
when to raise blame. This complexity is salient in the deterministic, sequential
version\footnote{cf Section 5.1 of \cite{KeilThiemannUnionIntersection}}. There,
the evaluation procedure has to deal again with the problem of eliding contracts
that are not \emph{compatible}\footnote{You can check more on this topic on
Section 5.1 of \cite{KeilThiemannUnionIntersection}} (the ones coming from a
different branch that shouldn't be applied in the current context). In practice,
the evaluation procedure maintains a global environment of contracts that is
extended each time a non-trivial contract (union, intersection or higher-order)
is opened. Then, any contract application must check the presence of the
corresponding label in the environment. But the most surprising\todo{(Yann) find
another word} part is that for contracts attached to a function that is applied,
the whole \emph{context} (the continuation of the application) must be checked
for compatibility with respect to each label in the contract
environment!\todo{(Yann) TBC: find better wording}

@@

% Another difficult issue Keil and Thiemann were able to solve, is the
% fact that user defined contracts and intersection do not usually play nice
% together, as shown on Section \ref{sec:flat-and-inter}.
% They provide two different sets of rules for their calculus,
% the first one is non deterministic, and its purpose is to provide
% semantics that are simple to understand.
% With these rules, the interaction between user defined contracts and
% intersections becomes much simpler, since they are able to check
% both sides of an intersection on separate threads.
% This allows them to be sure that contracts from one branch of the intersection
% won't get blamed while executing the other branch, and vice versa.
%
% As an example, consider the programs shown on Figure
% \ref{fig:kt-non-deterministic}, if the Original program
% gets executed, it will spawn two instances of said program,
% one with the first branch of the intersection (\nickel{3 | Num}),
% and the other one with the second branch (\nickel{3 | Positive}),
% and these will get joined once they reach an equal point, in this case
% it will be \nickel{3}.
%
% \begin{figure}[h]
% \begin{lstlisting}[language=nickel, title=Original]
% 3 | Num @& Positive
% \end{lstlisting}
% \begin{lstlisting}[language=nickel, title=First Program]
% 3 | Num
% \end{lstlisting}
% \begin{lstlisting}[language=nickel, title=Second Program]
% 3 | Positive
% \end{lstlisting}
% \begin{lstlisting}[language=nickel, title=Joined Program]
% 3
% \end{lstlisting}
% \caption{Non deterministic intersection contract opening}
% \label{fig:kt-non-deterministic}
% \end{figure}
%
% The second set of rules is deterministic, and it is intended
% to be simpler to implement.
% However, this second set needs to perform operations
% on the evaluation context, to decide whether or not to evaluate
% a given contract.
% \footnote{You can check more on this topic on Section 5.1 of
% \cite{KeilThiemannUnionIntersection}}
% For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
% when using the value \nickel{(fun x => x) | Str -> Str)}
% inside the execution of contract \nickel{C}, and applying it to \nickel{0},
% the contract runtime would check the context of execution and it would
% deduce that contract \nickel{Str -> Str} does not apply in this case,
% skipping it.
%
% We see two problems here, a minor one is the fact that the two
% different semantics defined for the calculus are really far apart
% one from the other, both on the strategies used for contract resolution,
% as well as on the complexity levels.
\resolved{(Yann) Is it really a
    problem per se? It is common thing to do (a nice declarative system
    for proofs and an algorithmic one for implementation, proved equivalent).
    Maybe the point is that the algorithmic system is way more complex than the
declarative one
(Teo) reworded}.
A second, major one,
is the dependency added over the evaluation context, since, again,
it makes composing this, with other features desired on programming languages,
difficult, or even impossible.
\resolved{(Yann) I think we need to substantiate this claim. If this
context dependency explains why CSE is invalid, maybe have a little example?
(Teo) I removed it, I don't think there's an example, mainly since
this context dependency is not part of the language. I do think it complicates
implementation of said feature}
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}
